---
pagetitle: "gpx"
editor_options: 
  chunk_output_type: console
---

<br>

## Working with `gpx` files

There are many file formats for spatial data. One of the more common formats used for many hand held devices (GPS, smart watches, etc.), `.gpx` are an interesting file type. They can hold points, as well as lines (or "tracks"). These usually have additional information, and include timestamps, elevation, and other metadata. 

**Let's load packages first:**

```{r loadlibsHidden, echo=TRUE, eval=FALSE, warning=FALSE}

library(sf)
library(dplyr)
library(lubridate)
library(mapview)
library(here)
mapviewOptions(fgb = FALSE)
```

```{r loadlibsOut, eval=TRUE, echo=FALSE, warning=FALSE}

suppressPackageStartupMessages({
  library(sf);
  library(dplyr);
  library(lubridate);
  library(mapview);
  library(here)
  }) 
mapviewOptions(fgb = FALSE)
```

### Read in a Track from `gpx` File

Let's use an example file that lives [here on github](https://raw.githubusercontent.com/ryanpeek/mapping-in-R-workshop/main/data/NFA.GPX) on github. Download the file, and save it into a `data` folder. 

Now we can read it in! Notice there's lots of additional information in there. However, this is a still a `data.frame` and an `sf` object, which means its easy to work with.

```{r gpxFile, echo=T, eval=T}

# a locally downloaded gpx file
file1 <- "NFA.GPX"

# read just the tracks:
trax <- st_read(here::here("data", file1), layer = "tracks")

# check the names of the tracks:
## here these represent dates for individual tracks
trax$name

# here we can see all the data inside the data frame
str(trax)

```

### Read points from a `gpx` from a URL

Now let's do the same thing, but let's read in `points` instead, and use a weblink instead of a local file. 

```{r gpxURL, echo=TRUE, eval=TRUE}

# read straight from the interwebs
file1url <- "https://raw.githubusercontent.com/ryanpeek/mapping-in-R-workshop/main/data/NFA.GPX"

# read just the tracks:
pts <- st_read(file1url, layer = "waypoints")

# check the names of attributes
names(pts)

```


Great!

### Pull a Single Track and Subset Waypoints

If we want to pull a single track or a single set of points, we can leverage `{dplyr}` and filter or subset our data in whatever way we want.

Let's grab a single track, and only pull waypoints from a selected set of locations and specific year (need the `{lubridate}` package).

```{r pullSingleTrack, echo=TRUE, eval=TRUE}

# pull a single track
trx1 <- trax %>% dplyr::filter(name=="2017-05-05-NFA")

# get points and exclude a site called "NFARRAV"

pts <- st_read(here("data/",file1), layer="waypoints") %>% 
	filter(!grepl("NFARRAV", x = name)) %>% 
	mutate(YYYY = year(time)) %>% 
  filter(YYYY == 2018)
```

### Mapview!

Now that we have our filtered and selected pieces, we can make a quick map with the excellent `{mapview}` package. We'll color points by elevation, just because. This information isn't particularly relevant, more to show how you can make fun mapview maps.

```{r mapview, eval=T, echo=T}

mapview(pts, zcol="ele") + mapview(trx1, color="orange")


```
