---
pagetitle: "mapview_and_leaflet"
---

## Using `mapview`

This package is pretty cool, and has a bunch of functionality. There are a number of [vignettes online already](https://r-spatial.github.io/mapview/articles/index.html), in particular check out the basic overview vignette [here](https://r-spatial.github.io/mapview/articles/articles/mapview_01-basics.html). 

### Main Functions

 - `mapview` - view potentially multiple spatial objects on a set of background maps
 - `viewExtent` - view extent / bounding box of spatial objects
 - `viewRGB` - view RGB true- or false-color images of raster objects
 - `latticeView` & `sync` - view small multiples of several potentially synchronised maps
 - `slideView` - view two overlaid raster objects with a horizontal slider to hide/reveal one of them
 - `cubeView` - 3D view of raster data cubes (e.g. space-time cube) with keyboard control over position of slices
 - `plainview` - view large rasters on a plain background (i.e. no background map)
 - `mapshot` - easily save maps (including leaflet maps) as html or png (or other image formats)

### Spatial Data Supported in `mapview`

It is natively set to work with `sf`, which is great. Note it will plot `raster` objects as well!

 - `sf` (XY dimension only)
 - `raster` (Layer, Stack, Brick and SpatialPixels* / SpatialGridDataFrame)
 - `sp` (Points, Polygons, Lines and their DataFrame version)
 - `satellite`

### Leaflet Compatible Functions

In addition, mapview provides some extra leaflet functionality:

 - `addHomeButton` - add zoom-to-layer buttons
 - `addLogo` - add a logo (or any image really) to a map
 - `addFeatures` - add features to a map (sf or sp). This is features type agnostic, meaning the same function can be used for points, lines and polygons
 - `addMouseCoordinates` - add mouse cursor postion information when moving over the map
 - `addLargeFeatures` - add large data to a map that would otherwise make the leaflet map irresponsive
 - `popupTable, popupGraph & popupImage` - popups of tables, grpahs and images for use with the leaflet add* functions

### Advanced `mapview` Options

https://r-spatial.github.io/mapview/articles/articles/mapview_02-advanced.html

## Using `leaflet`

You can also create `leaflet` maps, and they can be shared via html. This is a good way to provide dynamic spatial data, though bear in mind these html files can get large if you are trying to show a lot of data points or polygons.

### The Basics

The [`leaflet` website](https://rstudio.github.io/leaflet/) is excellent and should be the go-to for figuring out details. Below is a simple example building on some of the functions/data we worked with earlier in this workshop.

The main things to identify are:

 - What type of spatial data do you want to plot? (i.e., polygons, lines, markers, etc)
 - What format is that data? (works with `sf` or `sp` objects, or simple X/Y coordinates)
 - What base layers would you like to include? (see [here for the possibilities](http://leaflet-extras.github.io/leaflet-providers/preview/index.html))
 - How fancy do you want it to look? There are many extra options you can add using [built in plugins](https://rstudio.github.io/leaflet/morefeatures.html), or custom Javascript.

<hr>

## Example: California Snow Course Stations

Let's look at California snow course stations (check out the [CDEC website](http://cdec.water.ca.gov/snow/current/snow/index.html)). There's a nice package called "**`sharpshooter`**" we can use to pull the locations and metadata for all these stations and plot them with a `mapview` and `leaflet` map.

```{r GetData, echo=T, eval=T, purl=FALSE}

library(sharpshootR)  # CDEC.snow.courses, CDECquery, CDECsnowQuery

# GET DATA AND PREP 

data(CDEC.snow.courses)
snw<-CDEC.snow.courses

# make a few changes for plotting purposes
snw$id<-as.factor(snw$id)
snw$latitude<-as.numeric(snw$latitude)
snw$longitude<-as.numeric(snw$longitude)*-1
snw$apr1avg_in<-snw$april.1.Avg.inches
snw<-dplyr::select(snw, course_number, id, elev_feet:longitude,apr1avg_in)
str(snw) # check out data


```

### `mapview` Map

The nice thing is a mapview map is very simple to code compared to a leaflet map. See below.

```{r mapviewMap, echo=T, eval=T, purl=FALSE, warnings=FALSE, message=FALSE}

library(mapview)
library(sf)

# make a sf type object:
snw_sf <- st_as_sf(snw, coords = c("longitude", "latitude"), crs=4326, remove = FALSE)

# the default color scheme is viridis
mapview(snw_sf, zcol="apr1avg_in", layer.name="CDEC SNOW STATIONS")

```

<hr>

### `leaflet` Map

A leaflet map takes a bit more formatting, but the sky is the limit. We can make the map look pretty much exactly as we'd like, there are far fewer limitations.

```{r leafletMap, fig.width=7, echo=TRUE, eval=T, purl=F}

library(leaflet)


# add color palette
pal <- colorNumeric(
  palette = "GnBu",# can change to whatever: "RdBu", "GnBu"
  domain = snw$apr1avg_in
)

# Make a leaflet map!
m <- leaflet() %>% addTiles() %>% 
  #setView(lng = -120.8, lat = 39, zoom = 8) %>%  if you want to preset the view/zoom default
  addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") %>%
  addProviderTiles("Esri.WorldTopoMap", group = "Topo") %>%
  
  # add scale bar
  addMeasure(position = "topright",
             primaryLengthUnit = "meters",
             primaryAreaUnit = "sqmeters",
             activeColor = "#3D535D",
             completedColor = "#7D4479") %>%


# CDEC SNOW STATIONS
addCircleMarkers(data=snw, group="CDEC Snow",
                 lng= ~longitude, lat= ~latitude, # the spatial data, requires "~" here because it's NOT sp or sf object
                 popup=paste0("<strong>","Course ID: ","</strong>", 
                              snw$course_number, "<br><strong>", "Name: ",
                              "</strong>", snw$id, "<br><strong>", "Elev (ft): ",
                              "</strong>", snw$elev_feet, "<br><strong>", 
                              "Apr-1 Avg: ", "</strong>", snw$apr1avg_in),
                 stroke=TRUE, weight=0.6,radius=8,
                 fillOpacity = 0.5, color="black",
                 fillColor= ~pal(apr1avg_in)) %>% # mapping to the color palette 

  # add controls for basemaps and data
  addLayersControl(
    baseGroups = c("ESRI Aerial", "Topo"),
    overlayGroups = c("CDEC Snow"),
    options = layersControlOptions(collapsed = T))

m

```


